Abstractions and items: the measure noun phrase alternation in German

### 2017-12-25 Latex/article.tex
Nearing completion of Sec. 1 rewrite.

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer. Submitted April 2017. Do not quote.}
  %\affil[1]{}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{In this paper, a case alternation in German measure noun phrases is examined from the perspective of cognitively oriented corpus linguistics.
    In certain pseudo-partitive constructions, the embedded kind-denoting noun either agrees in its case with the measure head noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    My analysis is formulated in terms of Prototype Theory.
    I assume that the choice of alternants is influenced by non-alternating neighbouring constructions representing the prototypes more directly.
    I argue that the frequencies with which individual lemmas occur in these prototypical non-alternating constructions partially predict which alternant is chosen and that the genitive alternant is prototypical of strongly grammaticalised measure nouns.
    I present a large-scale corpus study using the DECOW corpus to support this theory.
    In the statistical analysis, I compare Maximum Likelihood estimators to Bayesian estimators (recently proposed for similar studies), showing that results \textit{predictably} do not differ.
    Finally, two experiments (forced choice and self-paced reading) are reported.
    The usage-based findings show a clear correlation with the behaviour of cognitive agents in both experiments.
    The present study therefore contributes to the well-established field of research into alternations using corpus and experimental methods.}
  \keywords{corpus methods and experimental methods, alternations, hierarchical models, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{Submitted April 2017. Do not quote.}
  \journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

\begin{acknowledgement}
  I thank (in alphabetical order) Felix Bildhauer, Susanne Flach, Elizabeth Pankratz, Samuel Reichert, Ulrike Sayatz, and Christian Zimmer for valuable discussions and comments.
  Also, I would like to thank two reviewers for Cognitive Linguistics with deep linguistics knowledge and knowledge of German grammar for comments which helped to improve the paper a lot.
  I thank Dagmar Divjak who, in her role as editor for Cognitive Linguistics, also read the paper thoroughly and provided very useful comments.

  Furthermore, I thank Ulrike Sayatz for helping me to recruit the participants for the experiments.
  Elizabeth Pankratz thankfully also fixed my English.
  Finally, I am grateful to my student assistants Kim Maser for her work on the annotation of the concordances and Luise Rißmann for supervising most of the experiments.
  The research presented here was made possible in part through funding from the \textit{Deutsche Forschungsgemeinschaft} (DFG, personal grant SCHA1916/1-1).
\end{acknowledgement}

>>>>

  \author*[1]{Roland Schäfer}
  \runningauthor{Roland Schäfer. Submitted April 2017. Do not quote.}
  %\affil[1]{}
  \affil[1]{Freie Universität Berlin}
  \title{Abstractions and items: the measure noun phrase alternation in German}
  \runningtitle{Abstraction and items: the measure noun phrase alternation German}
  %\subtitle{...}
  \abstract{In this paper, a case alternation in German measure noun phrases is examined from the perspective of cognitively oriented corpus linguistics.
    In certain pseudo-partitive constructions, the embedded kind-denoting noun either agrees in its case with the measure head noun (\textit{eine Tasse guter Kaffee} `a cup of good coffee') or it stands in the genitive (\textit{eine Tasse guten Kaffees}).
    My analysis is formulated in terms of Prototype Theory.
    I assume that the choice of alternants is influenced by non-alternating neighbouring constructions representing the prototypes more directly.
    I argue that the frequencies with which individual lemmas occur in these prototypical non-alternating constructions partially predict which alternant is chosen and that the genitive alternant is prototypical of strongly grammaticalised measure nouns.
    I present a large-scale corpus study using the DECOW corpus to support this theory.
    In the statistical analysis, I compare Maximum Likelihood estimators to Bayesian estimators (recently proposed for similar studies), showing that results \textit{predictably} do not differ.
    Finally, two experiments (forced choice and self-paced reading) are reported.
    The usage-based findings show a clear correlation with the behaviour of cognitive agents in both experiments.
    The present study therefore contributes to the well-established field of research into alternations using corpus and experimental methods.}
  \keywords{corpus methods and experimental methods, alternations, hierarchical models, pseudo-partitives, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{Submitted April 2017. Do not quote.}
  \journalyear{}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

\begin{acknowledgement}
  I thank (in alphabetical order) Felix Bildhauer, Susanne Flach, Elizabeth Pankratz, Samuel Reichert, Ulrike Sayatz, and Christian Zimmer for valuable discussions and comments.
  Also, I would like to thank the reviewers for Cognitive Linguistics as well as associate editor Dagmar Divjak for insightful comments which helped to improve the quality of this paper a lot.
  Furthermore, I thank Ulrike Sayatz for helping me to recruit the participants for the experiments.
  Elizabeth Pankratz thankfully also fixed my English.
  Finally, I am grateful to my student assistants Kim Maser for her work on the annotation of the concordances and Luise Rißmann for supervising most of the experiments.
  The research presented here was made possible in part through funding from the \textit{Deutsche Forschungsgemeinschaft} (DFG, personal grant SCHA1916/1-1).
\end{acknowledgement}

### 2017-12-25 Latex/sections/1intro.tex
Nearing completion of Sec. 1 rewrite.

This paper deals with a morpho-syntactic alternation between two constructions that occurs only in a very specific type of measure noun phrase in German.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.
The study of alternations has a long history in cognitively oriented corpus linguistics (for example, \citealp{BresnanEa2007,BresnanHay2010,BresnanFord2010,DivjakArppe2013,Gries2015,NessetJanda2010}).
This area of research is based on the assumption that language is a probabilistic phenomenon \citep{Bresnan2007} where alternants are chosen neither deterministically nor fully at random.
Instead, multifactorial models are constructed which incorporate influencing factors from diverse levels, including contextual factors.
The estimation of the model coefficients quantifies the influence that the factors have on the probability that either alternant is chosen.
There are two fundamental issues to consider.
First, there is a question of how corpus data relates to experimental findings, which provide more direct evidence of mental representations and processes.
Second, the appropriate modelling of such results in cognitive linguistics is a key issue.

As for the first issue, there has always been an interest in correlating probabilistic generalisations extracted from corpus data with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} rightly criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers.''
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.
It would be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted to \textit{test} the theory.
While the same approach is by and large applicable to corpus data.

\hl{Discussion of non-convergence here.}

\hl{Turning to the second issue, }
This entails that a variant of prototype theory with features is assumed \citep{Rosch1978}.
Prototype theory is well suited for modeling constructional choices, but it is just one of several similarity-based theories of classification, the most prominent other framework being exemplar theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in different experimental settings).
Prototype theory and exemplar theory model essentially the same types of effects but differ significantly in whether they assume higher-level abstractions in the form of single maximally prototypical exemplars or their features (prototype theory) or assume that categories emerge through the storage of many exemplars and similarity classification on those exemplars (exemplar theory).
\hl{Parallel to Langacker here}

As \cite{Barsalou1990} already showed, however, prototype and exemplar theory model the same types of effects and are informationally equivalent.
Consequently, experiments which favour one theory over the other use procedural behaviour of subjects, for example the speed of category retrieval, and not mere output data.
In very early experiments, \cite{PosnerKeele1968} showed, for example, that highly prototypical unseen exemplars were categorised more easily by subjects compared to less prototypical ones, even if these were included in the learning data.
Since corpus data only show artefacts of production events and we have no experimental access to the speaker's or writer's performance, one should be sceptical whether corpus analysis alone could ever decide which theory of mental representation is more suitable (see also \citealp[22]{Gries2003} and the \citealp[486--487]{Dabrowska2016} quote above).

>>>>

This paper deals with a morpho-syntactic alternation between two constructions that occurs only in a very specific type of measure noun phrase in German.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.
The study of constructional alternations has a long history in cognitively oriented corpus linguistics (for example, \citealp{BresnanEa2007,BresnanHay2010,BresnanFord2010,DivjakArppe2013,Gries2015,NessetJanda2010,Wulff2003}).
This area of research is based on the assumption that language is a probabilistic phenomenon \citep{Bresnan2007} where alternants are chosen neither deterministically nor fully at random.
Instead, multifactorial models are constructed which incorporate influencing factors from diverse levels, including contextual factors.
The estimation of the model coefficients quantifies the influence that the factors have on the probability that either alternant is chosen.
There are two fundamental issues to consider with respect to this tradition as a part of cognitive linguistics in the broad sense.
First, there is the question of how corpus data relates to experimental findings (which provide more direct evidence of mental representations and processes but are often intrinsically narrower in scope), and how well the two converge or should converge.
Second, the appropriate modelling of such results in cognitive linguistics is a key issue.

As for the first issue, there has always been an interest in correlating probabilistic generalisations extracted from corpus data with results from experimental work (for example, \citealp{ArppeJaervikivi2007,BresnanEa2007,BresnanFord2010,DivjakGries2008,DivjakEa2016,FordBresnan2013}).
This is often called a \textit{validation} of the corpus-derived findings, but \citet[303]{Divjak2016a} criticises this choice of words ``because it creates the impression that behavioral experimental data is inherently more valuable than textual data,'' citing \cite{TummersEa2005}, who state that a corpus is ``a sample of spontaneous language use that is (generally) realized by native speakers''.
However, as \citet[486--487]{Dabrowska2016} convincingly argues, this does not imply that we can in some way ``deduce mental representations from patterns of use,'' \ie\ from corpus data.%
\footnote{Epistemologically, I think she refers to acts of \textit{induction} rather than \textit{deduction}.}
It would be highly surprising if this were possible, and the same holds for experimental methods.
Nobody assumes that we can inductively infer mental representations from experiments, which -- as opposed to corpus studies -- even allow for direct access to the cognitive agent and offer much better possibilities to control experimental conditions and nuisance variables.
Rather, under the standard approach, a theory of cognitive representation is pre-specified.
Then, predictions are derived from this theory \textit{before} the experiment or the corpus study is conducted in order to \textit{test} the theory.
While the same approach is by and large applicable to corpus data.

An important question, however, is whether the usage data found in corpora are truly predictive of speakers' and writers' cognitive representation of language and\slash or of their overall linguistic behaviour.
An overview of this issue was given by \cite{NewmanSorensenduncan2015}, who enumerate a number of studies which show how corpus data and experimental data converge (such as \citealp{DurrantDoherty2010,BresnanEa2007,GriesWulff2005,GriesEa2005}) and a number of studies where the two types of data did not lead to diverging or only partially converging results (such as \citealp{ArppeJaervikivi2007,Dabrowska2014,Mollin2009}).
When researchers do not reach convergence, they mostly try to explain this by differentiating between the actual cognitive construct and what the pooled usage data as found in corpora represent, for example \citet[411]{Dabrowska2014}, who lists a long number of possible reasons to explain why subjects in her experiment diverged in their word association preferences from collocation measures extracted from corpora.
In some cases, researchers argue for a more adequate statistical analysis to increase the fit between corpus data and subjects' reactions in experiments, for example \cite{DivjakEa2016}, who show that generalised additive models (GAMs) are better suited than generalised linear models (GLMs) in the analysis of their reading-time experiment based on data extracted from previous corpus studies.
Much more optimistically compared to these publications dealing with possible frictions between usage and experimental data, \cite{StefanowitschFlach2016} recently proposed a straightforward positivist perspective of corpora as representing the input of an average adult speaker, thus licensing inferences from corpus data to cognitive representations under a ``corpus-as-input'' view.
They state that ``in this wider context, large, register-mixed corpora such as the British National Corpus [\ldots] may not be perfect models of the linguistic experience of adult speakers, but they are reasonably close to the input of an idealized average member of the relevant speech community'' \citep[104]{StefanowitschFlach2016}.%
\footnote{This, of course, raises the old question of how corpora should be composed to represent what they are intended to represent.
As points of reference for determining ``representativeness'', at least three major criteria have been used, namely the distribution of texts or text types in the output of all speakers of a language (production-based), the distribution of the relevance of texts or text types in the whole speech community (relevance-based), and the distribution of speakers' exposure to different texts or text types (perception-based).
See \citet{Biber1993}, \citet{MceneryEa2006}, \citet{Leech2007}, \citet{Hunston2008} for overviews from different perspectives.}
Clearly, no clear picture has yet emerged, which is not surprising given the vast number of cognitive constructs assumed at diverse levels (such as the level of words in collocation research and much more complex constructional levels in alternation research involving lexical, syntagmatic, and contextual factors), the problems of corpus composition, the operationalisations involved in making experiments, and in the choice of statistical tools.
While far from providing definitive solutions, my discussion in Section~\ref{sec:conclusion} will provide possible explanations for the goodness of fit between the corpus data and the experimental data reported in Sections~\ref{sec:corpusstudies} and~\ref{sec:experimental}, much in the spirit of \citet{Dabrowska2014}.

I now turn to the second issue, namely which type of cognitive representation research on alternations provides evidence for.
The typical approach in alternation research is to annotate a large number of corpus sentences with linguistic features and to model the probability of either variant being chosen given these features.
The idea is that a variant is chosen when the influencing features cumulatively assume \textit{prototypical} values for that variant.
I other words, a variant of prototype theory with features \citep{Rosch1978} suggests itself, and some researchers such as \cite{Gries2003}, \cite{NessetJanda2010}, or \cite{Schaefer2016c} indeed commit to prototype theory in alternation modelling.
Under prototype theory, category membership is defined by similarity to an ideal exemplar, which is usually identified with characterising features (see \citealp{Taylor2008} for an overview).
While prototype theory is well suited for modeling constructional choices, it is just one of at least two major similarity-based theories of classification, the most prominent other framework being exemplar theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in different experimental settings).
Prototype theory and exemplar theory model essentially the same types of effects but differ significantly in whether they assume higher-level abstractions in the form of single maximally prototypical exemplars or their features (prototype theory) or assume that categories emerge through the storage of many exemplars and similarity classification on those exemplars (exemplar theory).
\hl{Parallel to Langacker here}

As \cite{Barsalou1990} already showed, however, prototype and exemplar theory model the same types of effects and are informationally equivalent.
Consequently, experiments which favour one theory over the other use procedural behaviour of subjects in experiments, for example the speed of category retrieval, as opposed to mere output data.
In very early experiments, \cite{PosnerKeele1968} showed, for example, that highly prototypical unseen exemplars were categorised more easily by subjects compared to less prototypical ones, even if these were included in the learning data.
Since corpus data only show artefacts of production events and we have no experimental access to the speaker's or writer's performance, one should be sceptical whether corpus analysis alone could ever decide which theory of mental representation is more suitable (see also \citealp[22]{Gries2003} and the \citealp[486--487]{Dabrowska2016} quote above).
